Best Practices for Developing Monte Carlo Methodologies in Molecular Simulations [Article v\versionnumber]

### 2025-06-09 paper/paperv1.0.tex
Added quaternions back, fixed a few typos with delta theta max.

Orientations are typically fixed during particle translation trials so as not to introduce a second tunable parameter for orientation moves that cannot be decoupled from the translation parameter during optimization.
Instead, the orientation of a single particle may be perturbed in a separate trial move.
A robust and intuitive way to accomplish this is via the axis-angle method: (1) a particle is randomly chosen among $N_m$ mobile particles; (2) a rotation axis $\hat {\bf u}$ is generated by sampling uniformly on the surface of a sphere (Sec.~\ref{sec:unit_sphere}); (3) a rotation angle $\theta$ is generated uniformly on $\pm\delta\theta_{max}/2$ (with $\delta\theta_{max}$ a tunable parameter).
With the rotation so defined, a rotation matrix can be computed
\begin{equation}
\mathbf{R} =
\begin{bmatrix}
C + (1 - C) u_x^2 & (1 - C) u_x u_y - S u_z & (1 - C) u_x u_z + S u_y \\
(1 - C) u_y u_x + S u_z & C + (1 - C) u_y^2 & (1 - C) u_y u_z - S u_x \\
(1 - C) u_z u_x - S u_y & (1 - C) u_z u_y + S u_x & C + (1 - C) u_z^2
\end{bmatrix}
\label{eq:rotation_matrix}
\end{equation}
where $C =\cos\theta$ and $S = \sin\theta$.
The new coordinate of interaction site $j$ on particle $i$ at ${\bf r}_i$ is given by ${{\bf r}_{ij,{\rm n}}={\bf r}_{i}}+{\mathbf R}({\bf r}_{ij,{\rm o}}-{\bf r}_i)$.
If ${\bf r}_i$ is chosen as the centroid of the particle, then the centroid will be unchanged by the rotation (complementing the translation move, which leaves the orientation unchanged).
However, using the centroid for ${\bf r}_i$ is not a requirement; any well-defined center for the rotation may be employed (e.g., the position of a specific interaction site).

%Alternatively, if the orientation coordinate is held explicitly (e.g., as a quaternion), then it can be updated from $\hat{\bf u}$ and $\theta$ using an appropriate formula.

Regardless of the selected centroid, the transition probabilities are given in terms of the axis-angle selection process outlined above.
The elements of the transition probabilities for the particle orientation trial are summarized in Table~\ref{tab:lhs_rotation}.
Together they yield the forward and reverse probabilities, which in this case are again equal
\begin{equation}
  \alpha_{o\rightarrow n}=\alpha_{n\rightarrow o} = \frac{\diff\mathbf{\omega}\diff\theta}{N_m 4\pi\delta\theta_{max}}.
\label{eq:lhs_rhs_rot_fwdrev}
\end{equation}

%Instead, the orientation of a single particle may be perturbed in a separate trial move which proceeds as follows.
%A particle is randomly chosen among $N_m$ mobile particles.
%A new orientation of that particle is then chosen according to the probability $P_{\omega n}\diff\boldsymbol{\omega}$.
%The transition probabilities for the particle orientation trial are summarized in Table~\ref{tab:lhs_rotation}.

@@

@@

\begin{figure}
\lstinputlisting[
language=Python,
label={sim:gibbs},
caption=MC simulation of an ideal gas in the Gibbs ensemble with mechanical and chemical equilibrium using Python3 should yield two domains with equivalent number of particles and volume.
]{../codes/ideal_gas_gibbs.py}
\end{figure}

@@

\begin{figure}
\lstinputlisting[
language=Python,
label={sim:statistics},
caption=The statistics module computes the mean and the standard deviation of the mean using blocks.
]{../codes/statistics.py}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{\label{sec:unit_sphere}Random position on a unit sphere surface}
%%%%%%%%%%%%%%%%%%%%%%%%%

%In the example shown in Code Block~\ref{sim:unit_sphere}, positions are generated uniformly on the surface of a sphere.
%This is an important case to consider, because MC trials often assume that particles may be randomly oriented.
%However, one must be careful in how an orientation is generated.
%Uniformly selected values of Euler angles, for example, will result in a bias along the poles.
%Code Block~\ref{sim:quaternion} also shows how quaternions may be randomly generated to represent a uniformly random point on the surface of a sphere \cite{vesely_angular_1982}.
%Code Block~\ref{sim:rotationmatrix} uses these randomly generated quaternions to compute rotation matrices.
%A randomly generated perturbation of orientation is then demonstrated in Fig.~\ref{fig:random_orientation}.
Situations arise where one needs to generate a point at random uniformly on the surface of a sphere in 3D.  This is equivalent to defining an isotropically random direction or a randomly oriented unit vector. There are several ways to approach this: (1) sample the spherical coordinates $\theta$ and $\phi$, appropriately weighted; (2) use a rejection method to generate a point uniformly inside a sphere, i.e., repeatedly sample in a cube that circumscribes the sphere until the sample is also in the sphere, and then normalize the sampled point to a unit vector; (3) sample the $x$, $y$ and $z$ coordinates on independent Gaussian distributions, which yields a point samplied from a spherically symmetric 3D probability density, and normalize to a unit vector. The first method listed here is illustrated in Code Block~\ref{sim:unit_sphere}.

Code Block~\ref{sim:rotationmatrix} implements Eq.~\ref{eq:rotation_matrix} which are used to generate orientation-change trials in Code Block~\ref{sim:random_orientation} with the axis-angle method shown in Fig.~\ref{fig:random_orientation}.

\begin{figure}
\lstinputlisting[
language=Python,
label={sim:unit_sphere},
caption=The unit\_sphere module generates randomly uniform positions on the surface of a unit sphere in three dimensions.
]{../codes/unit_sphere.py}
\end{figure}

%\begin{figure}
%\lstinputlisting[
%language=Python,
%label={sim:quaternion},
%caption=The quaternion module generates a random quaternion about the surface of a 4-dimensional hypersphere.
%]{../codes/quaternion.py}
%\end{figure}

\begin{figure}
\lstinputlisting[
language=Python,
label={sim:rotationmatrix},
caption=The rotation\_matrix module generates a rotation matrix from an axis and angle.
]{../codes/rotation.py}
\end{figure}

\begin{figure}
\lstinputlisting[
language=Python,
label={sim:random_orientation},
caption=Visualize randomly uniform perturbations in three dimensional particle orientations.
]{../codes/random_orientation.py}
\end{figure}

\begin{figure}
\begin{centering}
\includegraphics[width=6cm]{../figures/random_orientation.png}
\caption{
The output from Code Block~\ref{sim:random_orientation} shows $10^3$ uniformly random blue unit vectors emanating from the origin.
The other $10^3$ orange unit vectors are confined to the top hemisphere because they were randomly generated with the axis-angle method by perturbing the orientation of an upwards $z$-axis vector (shown in black) with a maximum angle parameter, $\delta \theta_{max}=\pi/2=90$ degrees.
\label{fig:random_orientation}
}
\end{centering}
\end{figure}

@@

\begin{figure}
\lstinputlisting[
language=Python,
label={sim:spherical_shell},
caption=Visualize randomly uniform positions in a spherical shell in three dimensions.
]{../codes/spherical_shell.py}
\end{figure}

>>>>

Orientations are typically fixed during particle translation trials so as not to introduce a second tunable parameter for orientation moves that cannot be decoupled from the translation parameter during optimization.
Instead, the orientation of a single particle may be perturbed in a separate trial move.
A robust and intuitive way to accomplish this is via the axis-angle method: (1) a particle is randomly chosen among $N_m$ mobile particles; (2) a rotation axis $\hat {\bf u}$ is generated by sampling uniformly on the surface of a sphere (Sec.~\ref{sec:unit_sphere}); (3) a rotation angle $\theta$ is generated uniformly on $\pm\delta\theta_{max}/2$ (with $\delta\theta_{max}$ a tunable parameter).
With the rotation so defined, a rotation matrix can be computed
\begin{equation}
\mathbf{R} =
\begin{bmatrix}
C + (1 - C) u_x^2 & (1 - C) u_x u_y - S u_z & (1 - C) u_x u_z + S u_y \\
(1 - C) u_y u_x + S u_z & C + (1 - C) u_y^2 & (1 - C) u_y u_z - S u_x \\
(1 - C) u_z u_x - S u_y & (1 - C) u_z u_y + S u_x & C + (1 - C) u_z^2
\end{bmatrix}
\label{eq:rotation_matrix}
\end{equation}
where $C =\cos\theta$ and $S = \sin\theta$.
The new coordinate of interaction site $j$ on particle $i$ at ${\bf r}_i$ is given by ${{\bf r}_{ij,{\rm n}}={\bf r}_{i}}+{\mathbf R}({\bf r}_{ij,{\rm o}}-{\bf r}_i)$.
If the particle's centroid is chosen for ${\bf r}_i$, then the centroid will be unchanged by the rotation (complementing the translation move, which leaves the orientation unchanged).
However, using the centroid for ${\bf r}_i$ is not a requirement; any well-defined center for the rotation may be employed (e.g., the position of a specific interaction site).

Alternatively, if the orientation coordinate is held explicitly (e.g., as a quaternion), then it can be updated from $\hat{\bf u}$ and $\theta$ using an appropriate formula.

%Axis-angle is suitable for orientation displacement of particles of any symmetry. However, it is not suitable for generating a random orientation from scratch, i.e. by selecting $\delta = 2\pi$. Generating a random quaternion is best for this.
Regardless of the selected center for the rotation, the transition probabilities are given in terms of the axis-angle selection process outlined above.
The elements of the transition probabilities for the particle orientation trial are summarized in Table~\ref{tab:lhs_rotation}.
Together they yield the forward and reverse probabilities, which in this case are again equal
\begin{equation}
  \alpha_{o\rightarrow n}=\alpha_{n\rightarrow o} = \frac{\diff\mathbf{\omega}\diff\theta}{N_m 4\pi\delta\theta_{max}}.
\label{eq:lhs_rhs_rot_fwdrev}
\end{equation}

@@

If desired, generation of orientations from scratch, without perturbing on an existing orientation, is best accomplished by random generation of a quaternion.
An example of randomly picking quaternions, and thus orientations, is shown in Section~\ref{sec:unit_sphere}.

@@

\begin{figure}
\lstinputlisting[
language=Python,
label={sim:gibbs},
caption={MC simulation of an ideal gas in the Gibbs ensemble with mechanical and chemical equilibrium using Python3 should yield two domains with equivalent number of particles and volume.}
]{../codes/ideal_gas_gibbs.py}
\end{figure}

@@

\begin{figure}
\lstinputlisting[
language=Python,
label={sim:statistics},
caption={The statistics module computes the mean and the standard deviation of the mean using blocks.}
]{../codes/statistics.py}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{\label{sec:unit_sphere}Random position on a unit sphere surface and random orientation}
%%%%%%%%%%%%%%%%%%%%%%%%%

%In the example shown in Code Block~\ref{sim:unit_sphere}, positions are generated uniformly on the surface of a sphere.
%This is an important case to consider, because MC trials often assume that particles may be randomly oriented.
%However, one must be careful in how an orientation is generated.
%Uniformly selected values of Euler angles, for example, will result in a bias along the poles.
%Code Block~\ref{sim:quaternion} also shows how quaternions may be randomly generated to represent a uniformly random point on the surface of a sphere \cite{vesely_angular_1982}.
%Code Block~\ref{sim:rotationmatrix} uses these randomly generated quaternions to compute rotation matrices.
%A randomly generated perturbation of orientation is then demonstrated in Fig.~\ref{fig:random_orientation}.
Situations arise where one needs to generate a point at random uniformly on the surface of a sphere in 3D.  This is equivalent to defining an isotropically random direction or a randomly oriented unit vector. There are several ways to approach this: (1) sample the spherical coordinates $\theta$ and $\phi$, appropriately weighted; (2) use a rejection method to generate a point uniformly inside a sphere, i.e., repeatedly sample in a cube that circumscribes the sphere until the sample is also in the sphere, and then normalize the sampled point to a unit vector; (3) sample the $x$, $y$ and $z$ coordinates on independent Gaussian distributions, which yields a point sampled from a spherically symmetric 3D probability density, and normalize to a unit vector. The first method listed here is illustrated in Code Block~\ref{sim:unit_sphere}.

A similar problem is sampling a quaternion, which may be interpreted as a point on the surface of a unit 4D-hypersphere.
Accordingly, Code Block~\ref{sim:quaternion} shows how quaternions may be randomly generated by sampling uniformly on the surface of a 4D hypersphere using an efficient rejection method \cite{vesely_angular_1982}.
Code Block~\ref{sim:rotationmatrix} uses these randomly generated quaternions to compute rotation matrices, which could be used to generate a fully random orientation by rotating from an arbitrary reference orientation.
Finally, a randomly generated perturbation of orientation is shown using the axis-angle method in Code Block~\ref{sim:random_orientation} and Fig.~\ref{fig:random_orientation}.

\begin{figure}
\lstinputlisting[
language=Python,
label={sim:unit_sphere},
caption={The unit\_sphere module generates randomly uniform positions on the surface of a unit sphere in three dimensions.}
]{../codes/unit_sphere.py}
\end{figure}

\begin{figure}
\lstinputlisting[
language=Python,
label={sim:quaternion},
caption={The quaternion module generates a random quaternion by sampling on the surface of a 4-dimensional hypersphere.}
]{../codes/quaternion.py}
\end{figure}

\begin{figure}
\lstinputlisting[
language=Python,
label={sim:rotationmatrix},
caption={The rotation module generates a rotation matrix from an axis and angle or a quaternion. The first index is the row and the second index is the column, each starting with an index of zero.}
]{../codes/rotation.py}
\end{figure}

\begin{figure}
\lstinputlisting[
language=Python,
label={sim:random_orientation},
caption={Visualize randomly uniform perturbations in three dimensional particle orientations.}
]{../codes/random_orientation.py}
\end{figure}

\begin{figure}
\begin{centering}
\includegraphics[width=8.5cm]{../figures/random_orientation.png}
\caption{
The output from Code Block~\ref{sim:random_orientation} shows $10^3$ uniformly random blue unit vectors emanating from the origin.
The other $10^3$ orange unit vectors are confined to the top hemisphere because they were randomly generated with the axis-angle method by perturbing the orientation of an upwards $z$-axis vector (shown in black) with a maximum angle parameter, $\delta \theta_{max}=\pi=180$ degrees.
\label{fig:random_orientation}
}
\end{centering}
\end{figure}

@@

\begin{figure}
\lstinputlisting[
language=Python,
label={sim:spherical_shell},
caption={Visualize randomly uniform positions in a spherical shell in three dimensions.}
]{../codes/spherical_shell.py}
\end{figure}
